---
title: "Estimation of species or sites effects and interactions in abundance tables"
author: "Geneviève Robin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document we demonstrate how the package mimi can be used to analyse count data with covariates. 

## Aravo data

```{r load-aravo}
library(mimi)
library(ade4)
data(aravo)
```

We use an example from ecology with the Aravo data from the ade4 R package. It consists of an count table counting the abundance of 82 plant species across 75 environments in the French Alps. 

```{r aravo-counts}
aravo$spe[1:7, 1:5]
```

Covariates about environments (slope, snow, etc.) and species (height, mass, etc.) are also available.

```{r aravo-covariates}
aravo$env[1:7, 1:5]
aravo$traits[1:7, 1:5]
```



## Model

We use the methods in mimi with the following purpose: estimate the effects of the covariates on the counts, as well as additional interactions. Let $Y$ be the $n\times p$ count table, where $n$ is the number of environments and $p$ the number of species. Let $X_R$ be the $n\times q_R$ matrix containing the $q_R$ environments covariates, and $X_C$ be the $p\times q_C$ matrix containing the $q_C$ species covariates. 

We use the following extension of the log-bilinear model:
$$Y_{ij}\sim \mathcal{P}(\mu_{ij}), \quad \log(\mu_{ij}) = \mu + \langle x_{ij}|\alpha\rangle + \theta_{ij}, $$
where $x_{ij}$ is the vector containing the covariates for environment $i$ and species $j$: $x_{ij} = (X_R(i), X_C(j))$. Thus, $\mu$ is an offset, $\alpha$ is the vector of covariates effects, and $\theta_{ij}$ is an interaction term. \\

Of course, the model is overparametrized. We make the following assumptions: $\alpha$ is *sparse*, i.e. not all covariates influence the counts, and $\theta = (\theta_{ij})$ is low-rank, i.e. there are a few groups of similar species which can be summarized by a few latent variables. We estimate $\mu$, $\alpha$ and $\theta$ by minimizing the following penalized negative log-likelihood:

$$(\hat \mu, \hat\alpha, \hat\theta)\in \operatorname{argmin}\sum_{i=1}^n\sum_{j=1}^p (-Y_{ij}(\mu + \langle x_{ij}|\alpha\rangle + \theta_{ij}) + \exp(\mu + \langle x_{ij}|\alpha\rangle + \theta_{ij})) + \lambda_1\|\theta\|_{\star} + \lambda_2\|\alpha\|_1.$$
Here, the $\ell_1$ norm $\|\alpha\|_1$ (sum of absolute values of coefficients) and the nuclear norm $\|\theta\|_{\star}$ (sum of singular values) are convex relaxations of the sparsity and rank constraints respectively. The regularization parameters $\lambda_1$ and $\lambda_2$ act as trade-offs between fitting the data and privileging simpler models. They can be tuned with cross-validation for instance.

## Estimation of main effects and interactions with mimi

First of all we need to concatenate the matrices $X_R$ and $X_C$ to obtain a covariate vector $x_{ij}$ for every entry.

```{r create-covariate-matrix}
# Code the categorical variables using dummy variables
aravo$env$Form1 <- 1*(aravo$env$Form==1)
aravo$env$Form2 <- 1*(aravo$env$Form==2)
aravo$env$Form3 <- 1*(aravo$env$Form==3)
aravo$env <- aravo$env[,!(names(aravo$env)%in%"Form")]
aravo$env$ZoogDno <- 1*(aravo$env$ZoogD=="no")
aravo$env$ZoogDyes <- 1*(aravo$env$ZoogD=="yes")
aravo$env <- aravo$env[,!(names(aravo$env)%in%"ZoogD")]
X <- covmat(aravo$env, aravo$traits)
# scale and center the covariates
X <- scale(X)
```

Now we can proceed to the estimation

```{r mimi-estimation}
vt <- rep("poisson", ncol(aravo$spe)) # indicate the type of the columns of Y
res <- mimi(aravo$spe, model = "covariates", x = X, var.type = vt,
            lambda1 = 5, lambda2 = 1e-2)
df.alpha <- data.frame(var = c(colnames(aravo$env),
                               colnames(aravo$traits)),
                       alpha = res$list.alpha[[20]])
```

For a given value of $\lambda_1$ and $\lambda_2$ we obtain an estimate of the covariate effects and of the interaction. Here, the estimated interaction matrix is of rank $1$.
```{r results}
df.alpha
sum(svd(res$list.theta[[20]])$d>1e-6)

```

The parameters $\lambda_1$ and $\lambda_2$ can also be estimated by cross validation. We comment the code below because it is quite long to run.
```{r cv}
#res.cv <- cv.mimi(y = aravo$spe, model = "covariates", x = X)
```




